---
title: "An Analysis of the Community Reinvestment Act"
author: "Stephen Shannon, Swati Nayak, Alex Medeiros"
date: "2021-04-22"
output: ioslides_presentation
---



## Introduction 


  * Small businesses serve as the backbone of the economy, fostering innovation, job creation, and local development. 

  * However, access to financial resources remains a significant barrier to their growth, often influenced by the type of financial institutions and demographic factors.


  * This study examines the lending patterns of different banks to understand their role in supporting small businesses. 

  * Specifically, it focuses on **Community Banks (CBOs), Large Banks (LBOs), and Regional Banks (RBOs)**.

  * The goal is to analyze how these institutions distribute small business loans and whether certain banks provide more access than others.


## Importance

  * Additionally, this study investigates the impact of **demographic factors** on loan distribution and approval rates, including:

  * _Income levels_ – Do lower-income areas receive fewer loans?

  * _Race_ – Are certain racial groups disproportionately affected in loan approvals?

  * _Geographic location_ – How do urban and rural areas differ in accessing small business loans?


## Research Question


  * This study examines:

  * _Which types of banks engage in small business lending_ – Understanding the role of Community Banks (CBOs), Large Banks (LBOs), and Regional Banks (RBOs).

  * _How demographic factors influence loan distribution_ – Analyzing the impact of income, race, and geographic location on loan approvals.

  * _How these relationships have evolved over time_ – Examining trends in lending patterns and how they have changed due to economic and policy shifts.




## Key Findings?

  * Total Lending Trends:

  * Large Banks (LBOs) continue to dominate small business lending, issuing significantly higher loan amounts than Community Banks (CBOs) and Regional Banks (RBOs).

  * The relationship between loan amount and income is stronger in wealthier regions, meaning banks are more willing to lend to higher-income areas while underserved regions struggle with access to credit.

## Portfolio-Based Lending Differences:

  * LBOs lend in larger amounts but concentrate their lending in economically prosperous areas.

  * Over time, the proportion of loans from CBOs and RBOs has declined, further reducing financing options for small businesses in disadvantaged areas.


## The data


* CBO/LBO/RBO Definition:

  * CBOS: Less than 10b, RBOs 10b - 100b, LBOs: 100b+

* CRA (Community Reinvestment Act): A federal law enacted in 1977 to prevent discriminatory lending practices (redlining).

  * Timeseries, lending by bank to individual counties

* ACS (American Community Survey): A nationwide survey conducted by the U.S. Census Bureau to collect detailed demographic, economic, and housing data.

  * Demographic Information such as Household Income and population


```{r setup, include=FALSE, echo = FALSE, warn.conflicts =FALSE, results = 'hide'}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE, warn.conflicts =FALSE, results = 'hide'}

library(tidyverse)
library(data.table)
library(ggplot2)
library(plotly)

```

```{r, echo = FALSE, warn.conflicts = FALSE, results = 'hide', message=FALSE, warning=FALSE}

local_path <- "C:/Users/sshan/Desktop/umass amherst/sta633/stat_633/"

cra_data <- read.csv(paste0(local_path, "cra_data.csv"))

transmittal_data <- read.csv(paste0(local_path, "transmittal.csv")) %>% mutate(assets = assets)

acs_data <- read.csv(paste0(local_path,"acs.csv")) %>%
  mutate(FIPS = str_pad(FIPS, width = 5, pad = "0"))

cra_assets <- left_join(cra_data, transmittal_data, by = c("resp_id", "agency", "year")) %>%
  mutate(portfolio = case_when(
      # 10b CBO
    assets <= 10000000 ~ "CBO",
      # 10b - 100b RBO
    assets > 10000000 & assets <= 100000000 ~ "RBO",
      # 100b+ LBO
    assets > 100000000 ~ "LBO"
  )) %>% select(-1, -2)

county_data <- cra_assets %>%
  group_by(year, state, county, portfolio) %>%
  summarize(num_banks = n(),
            sum_loan_amt = sum(sum_loan_amt),
            sum_n_loans = sum(sum_n_loans)) %>% 
  mutate(state = str_pad(state, width = 2, pad = "0"),
         county = str_pad(county, width = 3, pad = "0")) %>% 
  unite(FIPS, state, county, sep = "")

county_acs_data <- left_join(county_data, acs_data, by = c("year" = "Year", "FIPS"))

```


## Plot 1


```{r, echo = FALSE, warn.conflicts = FALSE, message=FALSE, warning=FALSE}

# https://sshannon898.shinyapps.io/stat_633/

plot1_data <- cra_assets %>%
  filter(state <= 56) %>%
  mutate(state = str_pad(state, width = 2, pad = "0")) %>% 
  group_by(year, portfolio) %>%
  summarize(n = n(),
            amt_us_sb_loans = sum(sum_loan_amt),
            n_us_sb_loans = sum(sum_n_loans))

plot1 <- ggplot(data = plot1_data) +
  geom_area(aes(x = year,
            y = amt_us_sb_loans/1000000,
            fill = portfolio, group = portfolio)) +
  theme_bw(base_size = 16) +
  labs(x = "",
       y = "Total Lending (Billions $)")

plot1

```

Figure 1. The total small business lending amount by Bank type. 

 
## Plot 2 
 
```{r, echo = FALSE, warn.conflicts = FALSE, message=FALSE, warning=FALSE}


plot4 <- county_acs_data |>
  group_by(year, portfolio) |>
  summarize(mean_proportion = mean(sum_loan_amt / sum_n_loans, na.rm = TRUE)) |>
  ggplot() +
  geom_point(mapping = aes(x = year,
                           y = mean_proportion,
                           group = portfolio,
                           color = portfolio,
                           text = paste0("Year: ", year, "<br>Mean Loan Amount: ", round(mean_proportion), "<br>Portfolio: ", portfolio))) +
  geom_line(mapping = aes(x = year,
                          y = mean_proportion,
                          color = portfolio)) +
  labs(x = "",
       y = "Avg. Loan Size (Thousands)") + theme_bw(base_size = 16)
 
plot4


# ggplotly(plot4,
#          tooltip = "text")


```

Figure 2. Average Loan size by bank type. 


## Plot 3


```{r, echo = FALSE, warn.conflicts = FALSE, message=FALSE, warning=FALSE} 

# Load required packages

# library(RColorBrewer)


# 2. Calculate loan amount per capita
county_acs_data_temp <- county_acs_data %>%
  group_by(year) %>%
  mutate(loan_per_capita = sum_loan_amt / Population)

# 3. Classify counties into income categories using quantiles
# quantiles <- quantile(county_acs_data_temp$Median_HH_Income, probs = c(0.33, 0.66), na.rm = TRUE)
county_acs_data_temp <- county_acs_data_temp %>%
  filter(is.na(County) != TRUE) %>%
  group_by(year) %>%
  mutate(quantile_group = ntile(Median_HH_Income, 3)) %>%
  mutate(quantile_group = case_when(
    quantile_group == 1 ~ "Poor",
    quantile_group == 2 ~ "Middle",
    quantile_group == 3 ~ "Rich"
  ))

# 4. Group data by year and income category, then compute the average loan per capita
avg_loans <- county_acs_data_temp %>%
  group_by(year, portfolio, quantile_group) %>%
  summarise(avg_loan_amt = mean(sum_loan_amt, na.rm = TRUE))


p <- ggplot(data = avg_loans,
            aes(x = year, y = avg_loan_amt/1000, color = quantile_group, group = quantile_group)) + 
  geom_line() +
  theme_bw(base_size = 16) +
  labs(x = "",
       y = "Avg. Loan Size by County Income",
       color = "County Income") +
  facet_wrap(~portfolio) + scale_y_continuous()

p

```

Figure 3. Average Lending to counties of different income groups by bank size. 

## Shiny App

 Geographic Inequality in Lending:

  * Lending is concentrated in metropolitan and high-income regions, leaving rural areas and lower-income communities with fewer resources.

"https://sshannon898.shinyapps.io/stat_633/"


## Limitations

* This study does not include demographic variables, such as race, age, and sex.

* There is no data about loan approval and credit scores, which could certainly skew some of the data.

* Income across counties is very skewed, and this influences our results.

* Bank Categorizations are static across time, and a $10b threshold may not hold true in 2004 for a true "Community Bank"


## The end

Thank you!